---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

The code presumes that the starting data are downloaded into the working directory and unzipped. 
```{r, echo = TRUE, message=FALSE, tidy=TRUE}
act<-read.csv("activity.csv")
library(dplyr)
library(tidyr)
act <- tbl_df(act)
act_compl <- act[complete.cases(act),]
```

## What is mean total number of steps taken per day?

mean number of steps per day - **mean_steps**, median number of steps - **median_steps**
```{r, echo = TRUE, message=FALSE, tidy=TRUE}
act_date <- group_by(act_compl, date)
steps_per_day <- summarise_each(act_date, funs(sum))
hist(steps_per_day$steps, main="", xlab="Steps per day")
mean_steps <- mean(steps_per_day$steps)
median_steps <- median(steps_per_day$steps)
```
```{r, echo = TRUE, message=FALSE, tidy=TRUE}
print(mean_steps)
print(median_steps)
```

## What is the average daily activity pattern?

```{r, echo = TRUE, message=FALSE, tidy=TRUE}
act_int <- group_by(act_compl, interval)
act_int <- select(act_int, interval, steps)
act_int_mean <- summarise_each(act_int, funs(mean)) 
plot(act_int_mean$interval, act_int_mean$steps, type =  "l", main="Average daily activity pattern", xlab="Interval", ylab="Steps")
```

This fragment determines a 5-minute interval with maximum number of steps in average **inter_max**
```{r, echo = TRUE, message=FALSE, tidy=TRUE}
max_steps <- max(act_int_mean$steps)
max_int <- filter(act_int_mean, act_int_mean$steps == max_steps)
inter_max <- max_int$interval[[1]]
print(inter_max)
```

## Imputing missing values

```{r, echo = TRUE, message=FALSE, tidy=TRUE}
NA_number <- nrow(act) - nrow(act_compl)
print(NA_number)
```
This function looks if the element act[i, "steps"] is NA, if TRUE changes it to the value from the data table **act_int_mean**, which is a mean value of steps for that 5-minute interval.
```{r, echo = TRUE, message=FALSE, tidy=TRUE}
act_corr <- act
for (i in 1:nrow(act))
{if (is.na(act$steps[i]))
{int <- act$interval[i]
row <- filter(act_int_mean, act_int_mean$interval == int)
act_corr$steps[i] <- row$steps}}
act_corr_date <- group_by(act_corr, date)
steps_per_day_corr <- summarise_each(act_corr_date, funs(sum))
hist(steps_per_day_corr$steps, main="", xlab="Steps per day")
```

The histogramm differs from the previous one in the height of the "central bar" corresponding to the number of steps
that are close to the average.


## Are there differences in activity patterns between weekdays and weekends?

```{r, echo = TRUE, message=FALSE, tidy=TRUE}
day<-lapply(as.vector(act_corr$date), FUN=as.Date)
Sys.setlocale("LC_TIME","English_United States.1252")
wd<-sapply(day, FUN=weekdays)
for (i in 1:length(wd))
{if (wd[i]=="Saturday" | wd[i]=="Sunday")
wd[i]<-"weekend"
else wd[i]<-"weekday"}
act<-cbind(act_corr, wd)
act<-select(act, steps, interval, wd)
act<-group_by(act, wd, interval)
act_wd_mean <- summarise_each(act, funs(mean))
library(lattice)
xyplot(act_wd_mean$steps~act_wd_mean$interval|act_wd_mean$wd, 
       type="l",  layout=c(1,2),
       ylab="Number of steps", xlab="Interval")
```

There are some differences: e.g. on weekends there is no sharp peek in the morning and the activity is more even throughout the day.
